---
title: "Detection of non-adenosines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detection of non-adenosines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette describes the three analysis pipelines available in
**ninetails** for detecting non-adenosine residues in poly(A) tails.
For working with the output data, see `vignette("postprocessing")`.
For visualization, see `vignette("plotting")` and
`vignette("shiny_app")`.


## Dorado DRS Pipeline (Recommended)

`check_tails_dorado_DRS()` is the recommended function for direct RNA
sequencing (DRS) data basecalled with Dorado ≥ 1.0.0 using POD5 format.

```{r dorado-drs}
results <- ninetails::check_tails_dorado_DRS(
  dorado_summary = "path/to/dorado_alignment_summary.txt",
  pod5_dir       = "path/to/pod5_dir/",
  num_cores      = 2,
  qc             = TRUE,
  save_dir       = "~/output/",
  prefix         = "experiment1",
  part_size      = 40000,
  cleanup        = FALSE
)
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `dorado_summary` | character/data.frame | *required* | Path to Dorado summary file. Must contain: `read_id`, `filename`, `poly_tail_length`, `poly_tail_start`, `poly_tail_end` |
| `pod5_dir` | character | *required* | Path to directory containing POD5 files |
| `num_cores` | integer | `1` | Number of CPU cores for parallel processing |
| `qc` | logical | `TRUE` | Apply quality control filtering (see below) |
| `save_dir` | character | *required* | Output directory for results files |
| `prefix` | character | `""` | Optional prefix for output filenames (e.g., `"experiment1_"`) |
| `part_size` | integer | `40000` | Maximum reads per processing chunk (reduce if memory limited) |
| `cleanup` | logical | `FALSE` | Remove intermediate files after completion |

### Quality control (`qc = TRUE`)

When `qc = TRUE`, the pipeline applies the following filters before
signal analysis:

- Reads with `poly_tail_start = 0` are classified as `BAC` (bad
  coordinates) and excluded from signal processing
- Reads with `poly_tail_length < 10` are classified as `IRL`
  (insufficient read length)
- Unmapped reads (no `contig` assignment) are classified as `UNM`

These reads are still included in the output `read_classes` table but
are marked as `unclassified`.

### Generating the Dorado summary file

The Dorado summary file is produced by running `dorado summary` on your
aligned BAM file:

```bash
dorado summary aligned_reads.bam > dorado_summary.txt
```

The summary must include poly(A) tail estimates. Ensure that Dorado was
run with `--estimate-poly-a` during basecalling.

### Pipeline steps

The pipeline executes the following steps in order:

1. **Input validation** (`preprocess_inputs()`): Validates file paths,
   checks required columns, verifies POD5 directory structure
2. **Summary processing** (`process_dorado_summary()`): Reads and
   standardizes the Dorado summary file
3. **Quality filtering** (`filter_dorado_summary()`): Applies QC
   filters and assigns preliminary classifications
4. **Signal extraction** (`extract_tails_from_pod5()`): Extracts raw
   poly(A) tail signals from POD5 files via the Python `pod5` module
5. **Feature engineering** (`create_tail_features_list_dorado()`):
   Computes pseudomoves, signal statistics, and identifies candidate
   modification positions
6. **Segmentation** (`create_tail_chunk_list_dorado()`): Splits tail
   signals into 100-point segments centered on candidate positions
7. **GAF creation** (`process_dorado_signal_files()`): Converts segments
   to Gramian Angular Fields (GASF + GADF two-channel images)
8. **CNN classification** (`predict_gaf_classes()`): Applies the trained
   convolutional neural network to classify each segment as A, C, G, or
   U
9. **Output creation** (`create_outputs_dorado()`): Aggregates
   per-segment predictions into per-read classifications and produces
   final output tables

### Output files

Two tab-separated files are saved in `save_dir`:

- `{prefix}read_classes.txt` — one row per read
- `{prefix}nonadenosine_residues.txt` — one row per detected non-A
  residue (decorated reads only)

The function also returns both tables as a named list:

```{r dorado-output}
class_data <- results$read_classes
residue_data <- results$nonadenosine_residues
```


---

## Dorado cDNA Pipeline (Under Development)

> **Warning**
>
> This pipeline is currently under construction. Please do not use it
> for production analyses.

`check_tails_dorado_cDNA()` extends the DRS pipeline for complementary
DNA (cDNA) sequencing data with BAM file processing and automatic read
orientation classification.

```{r dorado-cdna}
results <- ninetails::check_tails_dorado_cDNA(
  bam_file       = "path/to/aligned_cdna.bam",
  dorado_summary = "path/to/dorado_summary.txt",
  pod5_dir       = "path/to/pod5_dir/",
  num_cores      = 2,
  qc             = TRUE,
  save_dir       = "~/output/",
  prefix         = "experiment1",
  part_size      = 40000,
  cleanup        = FALSE
)
```

### Additional parameter (vs DRS)

| Parameter | Type | Description |
|---|---|---|
| `bam_file` | character | Path to BAM file containing aligned cDNA reads with basecalled sequences |

All other parameters are identical to `check_tails_dorado_DRS()`.

### cDNA-specific processing

The cDNA pipeline includes additional steps not present in the DRS
pipeline:

1. **BAM splitting** (`split_bam_file_cdna()`): Splits the BAM file
   into manageable chunks
2. **Sequence extraction** (`extract_data_from_bam()`): Extracts
   basecalled sequences from BAM records
3. **Orientation detection** (`detect_orientation_single()` /
   `detect_orientation_multiple()`): Classifies each read as `polyA`,
   `polyT`, or `unidentified` using Dorado-style SSP/VNP primer
   matching against basecalled sequences
4. **Separate processing**: Poly(A) and poly(T) reads are processed
   independently through the signal analysis pipeline
5. **Result merging** (`merge_cdna_results()`): Combines poly(A) and
   poly(T) results into unified output tables

### Output

Output tables include an additional `tail_type` column (`polyA` or
`polyT`) indicating the read orientation.


---

## Guppy Legacy Pipeline

> **Warning**
>
> This pipeline works with Guppy basecaller 6.0.0 and lower only. It is
> maintained for backward compatibility and will not be further
> optimised. For new analyses, use the Dorado DRS pipeline.

`check_tails_guppy()` is the legacy function for DRS data basecalled
with Guppy using fast5 format and Nanopolish poly(A) coordinates.

```{r guppy-legacy}
results <- ninetails::check_tails_guppy(
  polya_data = system.file('extdata', 'test_data',
                           'nanopolish_output.tsv',
                           package = 'ninetails'),
  sequencing_summary = system.file('extdata', 'test_data',
                                   'sequencing_summary.txt',
                                   package = 'ninetails'),
  workspace = system.file('extdata', 'test_data',
                          'basecalled_fast5',
                          package = 'ninetails'),
  num_cores = 2,
  basecall_group = 'Basecall_1D_000',
  pass_only = TRUE,
  save_dir = '~/output/'
)
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `polya_data` | character/data.frame | *required* | Path to Nanopolish polya output or tailfindr output |
| `sequencing_summary` | character/data.frame | *required* | Path to sequencing summary file |
| `workspace` | character | *required* | Path to directory with multi-fast5 files |
| `num_cores` | integer | `1` | Number of CPU cores |
| `basecall_group` | character | `"Basecall_1D_000"` | Fast5 hierarchy level |
| `pass_only` | logical | `TRUE` | Include only "PASS" reads from Nanopolish QC |
| `qc` | logical | `TRUE` | Label terminal positions with "-WARN" suffix |
| `save_dir` | character | *required* | Output directory |
| `part_size` | integer | `1000000` | Max rows per processing chunk |

> **Note:** For tailfindR compatibility, see `vignette("tailfindr")`.


---

## Output structure

All pipelines return a named list with two data frames:

### read_classes

Complete accounting of all reads in the analysis.

| Column | Description |
|---|---|
| `readname` | Unique read identifier |
| `contig` | Reference sequence/transcript |
| `polya_length` | Estimated poly(A) tail length (nt) |
| `qc_tag` | Quality tag from basecaller QC |
| `class` | Classification: `decorated`, `blank`, or `unclassified` |
| `comments` | 3-letter code explaining the classification (see below) |
| `tail_type` | `polyA` or `polyT` (cDNA pipeline only) |

### nonadenosine_residues

Modification-level detail for `decorated` reads only. Each row
represents one detected non-adenosine residue.

| Column | Description |
|---|---|
| `readname` | Unique read identifier |
| `contig` | Reference sequence/transcript |
| `prediction` | Predicted nucleotide: `C`, `G`, or `U` |
| `est_nonA_pos` | Estimated position within the poly(A) tail (nucleotides from the 3' end) |
| `polya_length` | Total tail length (nt) |
| `qc_tag` | Quality tag |
| `tail_type` | `polyA` or `polyT` (cDNA pipeline only) |

> **Note:** A single read can have multiple rows in `nonadenosine_residues`
> if it contains more than one non-A residue.


---

## Classification codes

### Dorado pipelines (DRS and cDNA)

| Class | Code | Explanation |
|---|---|---|
| `decorated` | `YAY` | Non-adenosine residue detected |
| `blank` | `MAU` | No signal deviation; pure poly(A) tail |
| `blank` | `MPU` | Signal deviation present but predicted as adenosine |
| `unclassified` | `IRL` | Poly(A) tail too short (< 10 nt) |
| `unclassified` | `UNM` | Read unmapped |
| `unclassified` | `BAC` | Invalid coordinates (`poly_tail_start = 0`) |

### Guppy legacy pipeline

| Class | Code | Explanation |
|---|---|---|
| `decorated` | `YAY` | Move transition present, non-A detected |
| `blank` | `MAU` | Move transition absent, non-A undetected |
| `blank` | `MPU` | Move transition present, non-A undetected |
| `unclassified` | `QCF` | Nanopolish QC failed |
| `unclassified` | `NIN` | Not included (`pass_only = TRUE`) |
| `unclassified` | `IRL` | Insufficient read length |


---

## Memory considerations

> **Warning**
>
> Signal transformations can place a heavy load on memory, especially
> for full sequencing runs with millions of reads.

To manage memory:

- Adjust `part_size` to control reads per chunk. Default is 40,000
  (Dorado) or 1,000,000 (Guppy). Reduce to 10,000–20,000 if
  experiencing memory issues.
- Use `cleanup = TRUE` to remove intermediate files after processing
  completes.
- Monitor system memory during the GAF creation and CNN prediction
  steps, which are the most memory-intensive.
- For very large runs (>1M reads), consider splitting input files
  manually and processing in batches.


---

## Performance tips

- **Parallelization**: Set `num_cores` to the number of available CPU
  cores minus one. The pipeline parallelizes signal extraction, feature
  engineering, and GAF creation across chunks.
- **Disk space**: Intermediate files (signal data, feature lists, GAF
  matrices) can be large. Ensure sufficient free disk space in
  `save_dir`. Use `cleanup = TRUE` to remove intermediates automatically.
- **GPU acceleration**: If TensorFlow is configured with GPU support,
  the CNN classification step benefits from GPU acceleration. This is
  the single largest speedup available.
- **SSD storage**: Using SSD storage for `pod5_dir` and `save_dir`
  significantly reduces I/O bottlenecks during signal extraction.


---

## Next steps

After running the pipeline:

1. **Postprocess**: Load, annotate, merge, and summarize results —
   `vignette("postprocessing")`
2. **Visualize**: Generate static plots — `vignette("plotting")`
3. **Inspect signals**: Examine raw squiggles — `vignette("signal_inspection")`
4. **Explore interactively**: Launch the Shiny dashboard —
   `vignette("shiny_app")`
