---
title: "Detection of non-adenosines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detection of non-adenosines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette describes the three analysis pipelines available in **ninetails** for detecting non-adenosine residues in poly(A) tails.

## Dorado DRS Pipeline (Recommended)
 
`check_tails_dorado_DRS()` is the recommended function for direct RNA sequencing (DRS) data basecalled with Dorado â‰Ą 1.0.0 using POD5 format.

```{r dorado-drs}
results <- ninetails::check_tails_dorado_DRS(
  dorado_summary = "path/to/dorado_alignment_summary.txt",
  pod5_dir       = "path/to/pod5_dir/",
  num_cores      = 2,
  qc             = TRUE,
  save_dir       = "~/output/",
  prefix         = "experiment1",
  part_size      = 40000,
  cleanup        = FALSE
)
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `dorado_summary` | character/data.frame | *required* | Path to Dorado summary file. Must contain: `read_id`, `filename`, `poly_tail_length`, `poly_tail_start`, `poly_tail_end` |
| `pod5_dir` | character | *required* | Path to directory containing POD5 files |
| `num_cores` | integer | 1 | Number of CPU cores for parallel processing |
| `qc` | logical | TRUE | Apply quality control filtering |
| `save_dir` | character | *required* | Output directory |
| `prefix` | character | "" | Optional prefix for output filenames |
| `part_size` | integer | 40000 | Reads per processing chunk (reduce if memory limited) |
| `cleanup` | logical | FALSE | Remove intermediate files after completion |

### Pipeline steps

1. **Input Validation**: Validates Dorado summary files and POD5 directories
2. **Data Preprocessing**: Splits large datasets and applies quality filters
3. **Signal Extraction**: Extracts poly(A) tail signals from POD5 files
4. **Feature Engineering**: Computes pseudomoves and signal characteristics
5. **Segmentation**: Identifies candidate modification regions
6. **GAF Creation**: Creates Gramian Angular Fields for CNN input
7. **Classification**: Applies trained neural networks for prediction
8. **Output Creation**: Produces comprehensive results and statistics

---

## Dorado cDNA Pipeline (Under Development)

> **Warning**
>
> This pipeline is currently under construction. Please do not use it for production analyses.

`check_tails_dorado_cDNA()` extends the DRS pipeline for complementary DNA (cDNA) sequencing data with BAM file processing and automatic read orientation classification.

```{r dorado-cdna}
results <- ninetails::check_tails_dorado_cDNA(
  bam_file       = "path/to/aligned_cdna.bam",
  dorado_summary = "path/to/dorado_summary.txt",
  pod5_dir       = "path/to/pod5_dir/",
  num_cores      = 2,
  qc             = TRUE,
  save_dir       = "~/output/",
  prefix         = "experiment1",
  part_size      = 40000,
  cleanup        = FALSE
)
```

### Additional parameter (vs DRS)

| Parameter | Type | Description |
|---|---|---|
| `bam_file` | character | Path to BAM file containing aligned cDNA reads with basecalled sequences |

The cDNA pipeline classifies each read as `polyA`, `polyT`, or `unidentified` using Dorado-style SSP/VNP primer matching. Output tables include an additional `tail_type` column.

---

## Guppy Legacy Pipeline

> **Warning**
>
> This pipeline works with Guppy basecaller 6.0.0 and lower only. It is maintained for backward compatibility and will not be further optimised. For new analyses, use the Dorado DRS pipeline.

`check_tails_guppy()` is the legacy function for DRS data basecalled with Guppy using fast5 format and Nanopolish poly(A) coordinates.

```{r guppy-legacy}
results <- ninetails::check_tails_guppy(
  polya_data = system.file('extdata', 'test_data',
                           'nanopolish_output.tsv',
                           package = 'ninetails'),
  sequencing_summary = system.file('extdata', 'test_data',
                                   'sequencing_summary.txt',
                                   package = 'ninetails'),
  workspace = system.file('extdata', 'test_data',
                          'basecalled_fast5',
                          package = 'ninetails'),
  num_cores = 2,
  basecall_group = 'Basecall_1D_000',
  pass_only = TRUE,
  save_dir = '~/output/')
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `polya_data` | character/data.frame | *required* | Path to Nanopolish polya output or tailfindr output |
| `sequencing_summary` | character/data.frame | *required* | Path to sequencing summary file |
| `workspace` | character | *required* | Path to directory with multi-fast5 files |
| `num_cores` | integer | 1 | Number of CPU cores |
| `basecall_group` | character | "Basecall_1D_000" | Fast5 hierarchy level |
| `pass_only` | logical | TRUE | Include only "PASS" reads |
| `qc` | logical | TRUE | Label terminal positions with "-WARN" |
| `save_dir` | character | *required* | Output directory |
| `part_size` | integer | 1000000 | Max rows per chunk |

---

## Output structure

All pipelines return a named list with two data frames:

### read_classes

Complete accounting of all reads in the analysis.

| Column | Description |
|---|---|
| `readname` | Unique read identifier |
| `contig` | Reference sequence/transcript |
| `polya_length` | Estimated poly(A) tail length (nt) |
| `qc_tag` | Quality tag |
| `class` | Classification: `decorated`, `blank`, or `unclassified` |
| `comments` | 3-letter code explaining the classification |
| `tail_type` | `polyA` or `polyT` (cDNA pipeline only) |

### nonadenosine_residues

Modification-level detail for `decorated` reads only.

| Column | Description |
|---|---|
| `readname` | Unique read identifier |
| `contig` | Reference sequence/transcript |
| `prediction` | Predicted nucleotide: `C`, `G`, or `U` |
| `est_nonA_pos` | Position within poly(A) tail |
| `polya_length` | Total tail length (nt) |
| `qc_tag` | Quality tag |
| `tail_type` | `polyA` or `polyT` (cDNA pipeline only) |

---

## Classification codes

### Dorado pipelines (DRS and cDNA)

| Class | Code | Explanation |
|---|---|---|
| `decorated` | `YAY` | Non-adenosine residue detected |
| `blank` | `MAU` | No signal deviation; pure poly(A) |
| `blank` | `MPU` | Signal deviation present but predicted as adenosine |
| `unclassified` | `IRL` | Poly(A) tail too short (< 10 nt) |
| `unclassified` | `UNM` | Read unmapped |
| `unclassified` | `BAC` | Invalid coordinates (`poly_tail_start = 0`) |

### Guppy legacy pipeline

| Class | Code | Explanation |
|---|---|---|
| `decorated` | `YAY` | Move transition present, non-A detected |
| `blank` | `MAU` | Move transition absent, non-A undetected |
| `blank` | `MPU` | Move transition present, non-A undetected |
| `unclassified` | `QCF` | Nanopolish QC failed |
| `unclassified` | `NIN` | Not included (`pass_only = TRUE`) |
| `unclassified` | `IRL` | Insufficient read length |

---

## Memory considerations

> **Warning**
>
> Signal transformations can place a heavy load on memory, especially for full sequencing runs.

To manage memory:

- Adjust `part_size` to control reads per chunk
- Default is 40,000 (Dorado) or 1,000,000 (Guppy)
- Reduce if experiencing memory issues
- Use `cleanup = TRUE` to remove intermediate files
