---
title: "Get started with ninetails"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with ninetails}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

**Ninetails** detects and characterises non-adenosine nucleotides embedded within poly(A) tails of Oxford Nanopore sequencing reads. This vignette provides a quick introduction to installation and running the main analysis pipelines.

For complete documentation, see the [Ninetails Wiki](https://github.com/LRB-IIMCB/ninetails/wiki).

## Installation

**Ninetails** is not currently available on CRAN or Bioconductor. Install it from GitHub using `devtools`:

```{r installation}
install.packages("devtools")
devtools::install_github('LRB-IIMCB/ninetails')
library(ninetails)
```

> **Note for Windows users:**
> Before installing `devtools` on Windows, install Rtools so packages compile correctly: https://cran.r-project.org/bin/windows/Rtools/

### Additional dependencies

Depending on which pipeline you use, additional components are required:

**For Dorado pipelines (DRS and cDNA):**

- Python with `pod5` module: `pip install pod5`
- Keras/TensorFlow for R: `keras::install_keras()`

**For Guppy legacy pipeline:**

- `rhdf5` from Bioconductor: `BiocManager::install("rhdf5")`
- Keras/TensorFlow for R
- VBZ compression plugin (for newer MinKNOW data)

See the [Wiki](https://github.com/LRB-IIMCB/ninetails/wiki/3.-Additional-requirements) for detailed instructions.

## Available pipelines

Ninetails provides three analysis pipelines:

| Pipeline | Basecaller | Input format | Function | Status |
|---|---|---|---|---|
| **Dorado DRS** | Dorado ≥ 1.0.0 | POD5 + summary | `check_tails_dorado_DRS()` | Recommended |
| **Dorado cDNA** | Dorado ≥ 1.0.0 | POD5 + BAM + summary | `check_tails_dorado_cDNA()` | Under development |
| **Guppy** | Guppy ≤ 6.0.0 | fast5 + Nanopolish | `check_tails_guppy()` | Legacy |

## Quick start: Dorado DRS pipeline

The recommended pipeline for direct RNA sequencing (DRS) data:

```{r dorado-drs-quick}
results <- ninetails::check_tails_dorado_DRS(
  dorado_summary = "path/to/dorado_summary.txt",
  pod5_dir       = "path/to/pod5_dir/",
  num_cores      = 2,
  qc             = TRUE,
  save_dir       = "~/output/",
  prefix         = "my_experiment"
)
```

### Required inputs

- **dorado_summary**: Dorado summary file with columns `read_id`, `filename`, `poly_tail_length`, `poly_tail_start`, `poly_tail_end`
- **pod5_dir**: Directory containing POD5 files

### Output

The function returns a list with two data frames:

- **read_classes**: Per-read classification (decorated/blank/unclassified)
- **nonadenosine_residues**: Detailed non-A positions for decorated reads

Results are also saved as TSV files in `save_dir`.

## Next steps

- `vignette("detection")` — Detailed pipeline documentation
- `vignette("postprocessing")` — Working with results
- `vignette("plotting")` — Visualization functions
- [Ninetails Wiki](https://github.com/LRB-IIMCB/ninetails/wiki) — Complete documentation

## Citation

Please cite Ninetails as:

Gumińska, N., Matylla-Kulińska, K., Krawczyk, P.S. et al. Direct profiling of non-adenosines in poly(A) tails of endogenous and therapeutic mRNAs with Ninetails. *Nat Commun* **16**, 2664 (2025). https://doi.org/10.1038/s41467-025-57787-6
