---
title: "Get started with ninetails"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with ninetails}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

**Ninetails** detects and characterises non-adenosine nucleotides embedded within poly(A) tails of Oxford Nanopore sequencing reads. This vignette provides a quick introduction to running the main analysis pipelines.

For complete documentation including detailed installation instructions, data processing features, and visualisation tools, see the [Ninetails Wiki](https://github.com/LRB-IIMCB/ninetails/wiki).

## Installation

```{r installation}
install.packages("devtools")
devtools::install_github('LRB-IIMCB/ninetails')
library(ninetails)
```

**Note:** Ninetails requires additional external components (Dorado, POD5 Python module, Keras). See the [Wiki](https://github.com/LRB-IIMCB/ninetails/wiki) for full installation instructions.

## Available pipelines

Ninetails provides three analysis pipelines depending on your basecaller and data type:
                                                         
| Pipeline | Basecaller | Input format | Function |
|---|---|---|---|
| Dorado DRS | Dorado ≥ 1.0.0 | POD5 + summary | `check_tails_dorado_DRS()` |
| Dorado cDNA | Dorado ≥ 1.0.0 | POD5 + BAM + summary | `check_tails_dorado_cDNA()` |
| Guppy (legacy) | Guppy ≤ 6.0.0 | fast5 + Nanopolish | `check_tails_guppy()` |

## Dorado DRS pipeline (recommended)

The recommended pipeline for direct RNA sequencing (DRS) data basecalled with Dorado ≥ 1.0.0 using POD5 format.

```{r dorado-drs}
results <- ninetails::check_tails_dorado_DRS(
  dorado_summary = "path/to/dorado_alignment_summary.txt",
  pod5_dir       = "path/to/pod5_dir/",
  num_cores      = 2,
  qc             = TRUE,
  save_dir       = "~/output/",
  prefix         = "experiment1",
  part_size      = 40000,
  cleanup        = FALSE
)
```

### Key parameters

- `dorado_summary`: Path to Dorado summary file. Must contain `read_id`, `filename`, `poly_tail_length`, `poly_tail_start`, `poly_tail_end`.
- `pod5_dir`: Directory containing POD5 files.
- `num_cores`: Number of CPU cores for parallel processing.
- `qc`: Apply quality control filtering (recommended).
- `save_dir`: Output directory.
- `part_size`: Number of reads per processing chunk. Reduce if memory is limited.

## Dorado cDNA pipeline

For complementary DNA (cDNA) sequencing data. Extends the DRS pipeline with BAM file processing for sequence extraction and automatic read orientation classification (polyA vs polyT).

**Note:** This pipeline is currently under development.

```{r dorado-cdna}
results <- ninetails::check_tails_dorado_cDNA(
  bam_file       = "path/to/aligned_cdna.bam",
  dorado_summary = "path/to/dorado_summary.txt",
  pod5_dir       = "path/to/pod5_dir/",
  num_cores      = 2,
  qc             = TRUE,
  save_dir       = "~/output/",
  prefix         = "experiment1",
  part_size      = 40000,
  cleanup        = FALSE
)
```

The cDNA pipeline classifies each read as `polyA`, `polyT`, or `unidentified` using Dorado-style SSP/VNP primer matching before processing. Output tables include an additional `tail_type` column.

## Guppy legacy pipeline

For DRS data basecalled with Guppy ≤ 6.0.0 using fast5 format and Nanopolish poly(A) coordinates. **This pipeline is no longer actively developed** (critical bug fixes only).

```{r guppy-legacy}
results <- ninetails::check_tails_guppy(
  polya_data         = "path/to/nanopolish_output.tsv",
  sequencing_summary = "path/to/sequencing_summary.txt",
  workspace          = "path/to/basecalled_fast5/",
  num_cores          = 2,
  basecall_group     = 'Basecall_1D_000',
  pass_only          = TRUE,
  save_dir           = '~/output/'
)
```

## Output structure

All pipelines return a named list with two data frames:

### `read_classes`

Complete accounting of all reads. Each read is assigned one class and one comment code.

| class | comments | explanation |
|---|---|---|
| `decorated` | `YAY` | Non-adenosine residue detected |
| `blank` | `MAU` | No signal deviation detected; pure poly(A) signal |
| `blank` | `MPU` | Signal deviation present but predicted as adenosine only |
| `unclassified` | `IRL` | Poly(A) tail too short (< 10 nt) |
| `unclassified` | `UNM` | Read unmapped to reference |
| `unclassified` | `BAC` | Invalid poly(A) coordinates |

### `nonadenosine_residues`

Modification-level detail for `decorated` reads only. Columns include:

- `readname`: Unique read identifier
- `contig`: Reference sequence/transcript
- `prediction`: Predicted nucleotide type (`C`, `G`, or `U`)
- `est_nonA_pos`: Estimated position within the poly(A) tail from the 3' end
- `polya_length`: Total poly(A) tail length (nt)

## Further resources

- [Ninetails Wiki](https://github.com/LRB-IIMCB/ninetails/wiki) — full documentation
- [GitHub Issues](https://github.com/LRB-IIMCB/ninetails/issues) — bug reports and feature requests
- [Publications](publications.html) — papers using ninetails
