---
title: "Get started with ninetails"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with ninetails}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

**Ninetails** detects and characterises non-adenosine nucleotides
embedded within poly(A) tails of Oxford Nanopore sequencing reads. It
uses convolutional neural networks applied to Gramian Angular Field
representations of raw current signals to identify cytidine (C),
guanosine (G), and uridine (U) residues within otherwise pure poly(A)
tails.

For complete documentation, see the
[Ninetails Wiki](https://github.com/LRB-IIMCB/ninetails/wiki).


## Installation

**Ninetails** is not currently available on CRAN or Bioconductor.
Install it from GitHub using `devtools`:

```{r installation}
install.packages("devtools")
devtools::install_github('LRB-IIMCB/ninetails')
library(ninetails)
```

> **Note for Windows users:**
> Before installing `devtools` on Windows, install Rtools so packages
> compile correctly: https://cran.r-project.org/bin/windows/Rtools/

### Additional dependencies

Depending on which pipeline you use, additional components are required:

**For Dorado pipelines (DRS and cDNA):**

- Python with `pod5` module: `pip install pod5`
- Keras/TensorFlow for R: `keras::install_keras()`
- The `reticulate` R package for Python interoperability

**For Guppy legacy pipeline:**

- `rhdf5` from Bioconductor: `BiocManager::install("rhdf5")`
- Keras/TensorFlow for R
- VBZ compression plugin (for newer MinKNOW data)

**For the interactive dashboard** (optional):

- `shiny`, `plotly`, `htmltools`, `DT`, `base64enc`, `cowplot`
- Install with:
  `install.packages(c("shiny", "plotly", "htmltools", "DT", "base64enc", "cowplot"))`

See the
[Wiki](https://github.com/LRB-IIMCB/ninetails/wiki/3.-Additional-requirements)
for detailed instructions.


## Available pipelines

Ninetails provides three analysis pipelines:

| Pipeline | Basecaller | Input format | Function | Status |
|---|---|---|---|---|
| **Dorado DRS** | Dorado ≥ 1.0.0 | POD5 + summary | `check_tails_dorado_DRS()` | Recommended |
| **Dorado cDNA** | Dorado ≥ 1.0.0 | POD5 + BAM + summary | `check_tails_dorado_cDNA()` | Under development |
| **Guppy** | Guppy ≤ 6.0.0 | fast5 + Nanopolish | `check_tails_guppy()` | Legacy |

Choose the pipeline that matches your basecaller and sequencing protocol.
The Dorado DRS pipeline is recommended for all new analyses.


## Quick start: Dorado DRS pipeline

The recommended pipeline for direct RNA sequencing (DRS) data:

```{r dorado-drs-quick}
results <- ninetails::check_tails_dorado_DRS(
  dorado_summary = "path/to/dorado_summary.txt",
  pod5_dir       = "path/to/pod5_dir/",
  num_cores      = 2,
  qc             = TRUE,
  save_dir       = "~/output/",
  prefix         = "my_experiment"
)
```

### Required inputs

- **dorado_summary**: Dorado summary file with columns `read_id`,
  `filename`, `poly_tail_length`, `poly_tail_start`, `poly_tail_end`.
  Generate this with `dorado summary` on your aligned BAM file.
- **pod5_dir**: Directory containing POD5 files from the sequencing run.

### Output

The function returns a named list with two data frames:

- **read_classes**: Per-read classification (decorated, blank, or
  unclassified) with columns `readname`, `contig`, `polya_length`,
  `qc_tag`, `class`, and `comments`.
- **nonadenosine_residues**: Detailed non-A positions for decorated reads
  with columns `readname`, `contig`, `prediction` (C/G/U),
  `est_nonA_pos`, `polya_length`, and `qc_tag`.

Results are also saved as tab-separated files in `save_dir`.

### Classification codes

| Class | Code | Meaning |
|---|---|---|
| `decorated` | `YAY` | Non-adenosine residue detected |
| `blank` | `MAU` | No signal deviation; pure poly(A) |
| `blank` | `MPU` | Signal deviation present but predicted as adenosine |
| `unclassified` | `IRL` | Poly(A) tail too short (< 10 nt) |
| `unclassified` | `UNM` | Read unmapped |
| `unclassified` | `BAC` | Invalid coordinates |


## Working with results

After running the pipeline, use the postprocessing functions to load,
merge, annotate, and summarize the data:

```{r postprocessing}
# Load multiple samples
class_data <- ninetails::read_class_multiple(samples_table)
residue_data <- ninetails::read_residue_multiple(samples_table)

# Merge into one table
merged <- ninetails::merge_nonA_tables(class_data, residue_data)

# Annotate with gene symbols (requires biomaRt)
class_data <- ninetails::annotate_with_biomart(class_data, species = "mmusculus")
residue_data <- ninetails::annotate_with_biomart(residue_data, species = "mmusculus")

# Summarize per transcript
summary <- ninetails::summarize_nonA(merged, summary_factors = "group")
```

See `vignette("postprocessing")` for the complete workflow.


## Visualization

Ninetails provides both static plotting functions and an interactive
Shiny dashboard.

### Static plots

```{r plotting}
# Read classification
ninetails::plot_class_counts(class_data, grouping_factor = "sample_name")

# Residue frequency
ninetails::plot_residue_counts(residue_data, grouping_factor = "sample_name")

# Non-A abundance per read
ninetails::plot_nonA_abundance(residue_data, grouping_factor = "sample_name")

# Poly(A) tail length
ninetails::plot_tail_distribution(class_data, grouping_factor = "sample_name")

# Non-A position distribution
ninetails::plot_rug_density(residue_data, base = "C", max_length = 200)
```

See `vignette("plotting")` for all plotting functions and parameters.

### Interactive dashboard

Launch the Shiny dashboard for interactive exploration of results with
configurable filters, per-sample rug plots, signal visualization, and
downloadable reports:

```{r dashboard}
ninetails::launch_signal_browser(
  class_file   = "/path/to/read_classes.txt",
  residue_file = "/path/to/nonadenosine_residues.txt",
  summary_file = "/path/to/dorado_summary.txt",
  pod5_dir     = "/path/to/pod5/"
)
```

See `vignette("shiny_app")` for complete dashboard documentation.


## Next steps

- `vignette("detection")` — Detailed pipeline documentation and
  parameters
- `vignette("postprocessing")` — Loading, merging, correcting, and
  summarizing results
- `vignette("plotting")` — All visualization functions with parameter
  tables
- `vignette("signal_inspection")` — Raw signal visualization from
  fast5 and POD5 files
- `vignette("shiny_app")` — Interactive analysis dashboard
- [Ninetails Wiki](https://github.com/LRB-IIMCB/ninetails/wiki) —
  Complete documentation


## Troubleshooting

**Keras/TensorFlow not found**
Run `keras::install_keras()` to install the Python backend. Ninetails
requires a working Keras installation for the CNN classification step.

**POD5 module not available**
Install the Python pod5 package: `pip install pod5`. Verify with
`reticulate::py_module_available("pod5")`.

**Memory issues during processing**
Reduce the `part_size` parameter (default: 40,000 reads per chunk).
Use `cleanup = TRUE` to remove intermediate files automatically.

**Empty results**
Check that your Dorado summary file contains the required columns:
`read_id`, `filename`, `poly_tail_length`, `poly_tail_start`,
`poly_tail_end`. Reads with `poly_tail_start = 0` are classified as
`BAC` (bad coordinates).


## Citation

Please cite Ninetails as:

Gumińska, N., Matylla-Kulińska, K., Krawczyk, P.S. et al. Direct
profiling of non-adenosines in poly(A) tails of endogenous and
therapeutic mRNAs with Ninetails. *Nat Commun* **16**, 2664 (2025).
https://doi.org/10.1038/s41467-025-57787-6
