---
title: "Data postprocessing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data postprocessing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette describes functions for loading, processing, annotating,
and analyzing **ninetails** output data. All functions documented here
operate on the two output tables produced by the main pipelines
(`check_tails_dorado_DRS()`, `check_tails_dorado_cDNA()`, or
`check_tails_guppy()`):

- **read_classes**: Per-read classification (decorated/blank/unclassified)
- **nonadenosine_residues**: Modification-level detail for decorated reads

For visualization of these data, see `vignette("plotting")` and
`vignette("shiny_app")`.


## Reading files into R

### Reading single files

```{r read-single}
# Read read_classes output
class_path <- "/path/to/read_classes.tsv"
class_data <- ninetails::read_class_single(class_path)

# Read nonadenosine_residues output
residue_path <- "/path/to/nonadenosine_residues.tsv"
residue_data <- ninetails::read_residue_single(residue_path)
```

| Function | Input | Output |
|---|---|---|
| `read_class_single()` | Path to a single `read_classes` file | Data frame with `readname`, `contig`, `polya_length`, `qc_tag`, `class`, `comments` |
| `read_residue_single()` | Path to a single `nonadenosine_residues` file | Data frame with `readname`, `contig`, `prediction`, `est_nonA_pos`, `polya_length`, `qc_tag` |


### Reading multiple files

For experiments with multiple samples, use `read_class_multiple()` and
`read_residue_multiple()` with a metadata table. The metadata table must
contain columns `sample_name`, `group`, `class_path`, and
`residue_path`.

```{r read-multiple}
# Define metadata table
samples_table <- data.frame(
  sample_name = c("sample_1", "sample_2", "sample_3", "sample_4"),
  group = c("control", "control", "treatment", "treatment"),
  class_path = c(
    "/path/to/sample_1/read_classes.tsv",
    "/path/to/sample_2/read_classes.tsv",
    "/path/to/sample_3/read_classes.tsv",
    "/path/to/sample_4/read_classes.tsv"
  ),
  residue_path = c(
    "/path/to/sample_1/nonadenosine_residues.tsv",
    "/path/to/sample_2/nonadenosine_residues.tsv",
    "/path/to/sample_3/nonadenosine_residues.tsv",
    "/path/to/sample_4/nonadenosine_residues.tsv"
  )
)

# Read all files at once
class_data <- ninetails::read_class_multiple(samples_table)
residue_data <- ninetails::read_residue_multiple(samples_table)
```

The output data frames include additional `sample_name` and `group`
columns (as factors), which are used as grouping variables in downstream
analysis and plotting.


### Using a YAML configuration file

Alternatively, provide metadata in a YAML configuration file. This is
the same format used by the Shiny dashboard
(see `vignette("shiny_app")`):

```{r read-yaml}
config <- yaml::yaml.load_file("config.yml")
samples_table <- data.frame(t(sapply(config$samples, unlist)))
rownames(samples_table) <- NULL

class_data <- ninetails::read_class_multiple(samples_table)
residue_data <- ninetails::read_residue_multiple(samples_table)
```

Example `config.yml`:

```yaml
samples:
  sample_1:
    sample_name: sample_1
    group: control
    class_path: /path/to/sample_1/read_classes.tsv
    residue_path: /path/to/sample_1/nonadenosine_residues.tsv
  sample_2:
    sample_name: sample_2
    group: treatment
    class_path: /path/to/sample_2/read_classes.tsv
    residue_path: /path/to/sample_2/nonadenosine_residues.tsv
```


---

## Annotation with biomaRt

`annotate_with_biomart()` adds gene symbols and Ensembl transcript IDs
to ninetails output tables by querying biomaRt. This enables
transcript-level filtering by human-readable gene names instead of raw
contig identifiers.

```{r annotate}
# Annotate class data
class_data <- ninetails::annotate_with_biomart(
  class_data,
  species = "mmusculus"
)

# Annotate residue data
residue_data <- ninetails::annotate_with_biomart(
  residue_data,
  species = "mmusculus"
)
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `data` | data.frame | *required* | Ninetails output table with a `contig` column |
| `species` | character | *required* | Species identifier (see below) |

### Supported species

| Identifier | Species |
|---|---|
| `"mmusculus"` | *Mus musculus* |
| `"hsapiens"` | *Homo sapiens* |
| `"athaliana"` | *Arabidopsis thaliana* |
| `"scerevisiae"` | *Saccharomyces cerevisiae* |
| `"celegans"` | *Caenorhabditis elegans* |
| `"tbrucei"` | *Trypanosoma brucei* |

See function documentation for other species and custom biomaRt datasets.

### Output columns added

| Column | Description |
|---|---|
| `symbol` | Gene symbol (e.g., `Actb`, `Gapdh`) |
| `ensembl_transcript_id_short` | Ensembl transcript ID without version suffix |

> **Note:** Annotation should be performed before merging, summarizing,
> or launching the dashboard, since downstream functions and the
> dashboard use the `symbol` column for transcript labels when available.


---

## Correcting classification

Sometimes nucleotides from AT-rich transcript ends are misidentified as
poly(A) tails. Use `reclassify_ninetails_data()` to minimize
segmentation artifacts by comparing detected residue positions against
known transcript sequences:

```{r reclassify}
ninetails_data <- ninetails::reclassify_ninetails_data(
  residue_data      = residue_data,
  class_data        = class_data,
  grouping_factor   = "sample_name",
  transcript_column = "ensembl_transcript_id_short",
  ref               = "mmusculus"
)

# Retrieve corrected data frames
class_data <- ninetails_data[[1]]
residue_data <- ninetails_data[[2]]
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `residue_data` | data.frame | *required* | Non-A residue table |
| `class_data` | data.frame | *required* | Read classification table |
| `grouping_factor` | character | *required* | Grouping column (e.g., `"sample_name"`) |
| `transcript_column` | character | *required* | Column with transcript IDs |
| `ref` | character | *required* | Reference species (same identifiers as `annotate_with_biomart()`) |

> **Note:** Apply this function before further analysis. It requires
> annotated data (i.e., run `annotate_with_biomart()` first).


---

## Merging results

Combine class and residue data into a single table with one row per
read. Decorated reads gain a `nonA_residues` column summarizing all
detected positions.

```{r merge}
merged_tables <- ninetails::merge_nonA_tables(
  class_data  = class_data,
  residue_data = residue_data,
  pass_only   = TRUE
)
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `class_data` | data.frame | *required* | Read classification table |
| `residue_data` | data.frame | *required* | Non-A residue table |
| `pass_only` | logical | `TRUE` | If `TRUE`, include only QC-passed reads |

### Output

The merged table contains all columns from `class_data` plus a
`nonA_residues` column. For decorated reads, this column contains all
non-A positions from 5' to 3' separated by `:` (e.g., `"C:G"` for a
read with a cytidine and a guanosine). For blank and unclassified reads,
the column is `NA`.


### Spreading residue columns

`spread_nonA_residues()` reshapes the merged table to create separate
columns for each non-A position, which can be useful for per-position
analysis:

```{r spread}
spread_table <- ninetails::spread_nonA_residues(
  merged_nonA_tables = merged_tables
)
```

The output adds columns `nonA_1`, `nonA_2`, etc., each containing
the residue type at that position (or `NA` if the read has fewer
modifications).


---

## Counting and summarizing

### Count by class

```{r count-class}
# Simple counts
class_counts <- ninetails::count_class(class_data)

# Grouped by sample
class_counts <- ninetails::count_class(
  class_data,
  grouping_factor = "sample_name"
)

# Detailed (by comments code) vs simplified (by class)
counts_detailed <- ninetails::count_class(class_data, detailed = TRUE)
counts_simple <- ninetails::count_class(class_data, detailed = FALSE)
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `class_data` | data.frame | *required* | Read classification table |
| `grouping_factor` | character | `NA` | Grouping column |
| `detailed` | logical | `TRUE` | If `TRUE`, count by comment code; if `FALSE`, count by class |


### Count residues

```{r count-residues}
residue_counts <- ninetails::count_residues(
  residue_data,
  grouping_factor = "sample_name"
)
```

### Count non-A abundance

Count how many reads have one, two, or three or more non-A residues:

```{r count-abundance}
abundance <- ninetails::count_nonA_abundance(
  residue_data,
  grouping_factor = "sample_name"
)
```


---

## Summarizing by transcript

Generate comprehensive summary statistics per transcript:

```{r summarize}
summarized <- ninetails::summarize_nonA(
  merged_nonA_tables    = merged_tables,
  summary_factors       = "group",
  transcript_id_column  = "ensembl_transcript_id_short"
)
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `merged_nonA_tables` | data.frame | *required* | Output from `merge_nonA_tables()` |
| `summary_factors` | character | *required* | Grouping column(s) for the summary |
| `transcript_id_column` | character | `NULL` | Column with transcript IDs (optional) |

### Output columns

| Column | Description |
|---|---|
| Grouping columns | As specified by `summary_factors` |
| `contig` / transcript ID | Transcript identifier |
| `n_reads` | Total reads mapping to the transcript |
| `n_decorated` | Number of decorated reads |
| `n_blank` | Number of blank reads |
| `n_C_counts` / `n_G_counts` / `n_U_counts` | Reads containing C, G, or U residues |
| `n_C_hits` / `n_G_hits` / `n_U_hits` | Total C, G, or U residue occurrences |
| `mean_polya_length` | Mean poly(A) tail length |
| `median_polya_length` | Median poly(A) tail length |


---

## Statistical analysis

### Fisher's exact test

Compare non-adenosine frequencies between two conditions:

```{r fisher}
fisher_results <- ninetails::calculate_fisher(
  ninetails_data,
  grouping_factor = "condition",
  comparison = c("control", "treatment")
)
```

### Parameters

| Parameter | Type | Description |
|---|---|---|
| `ninetails_data` | data.frame | Class data or merged table |
| `grouping_factor` | character | Column defining conditions |
| `comparison` | character(2) | Two condition names to compare |

The function returns a data frame with Fisher's exact test results
(p-value, odds ratio, confidence interval) per transcript.


---

## Typical workflow

```{r workflow}
library(ninetails)

# 1. Load data from multiple samples
samples <- data.frame(
  sample_name = c("WT_1", "WT_2", "KO_1", "KO_2"),
  group = c("WT", "WT", "KO", "KO"),
  class_path = c("WT_1_classes.tsv", "WT_2_classes.tsv",
                 "KO_1_classes.tsv", "KO_2_classes.tsv"),
  residue_path = c("WT_1_residues.tsv", "WT_2_residues.tsv",
                   "KO_1_residues.tsv", "KO_2_residues.tsv")
)

class_data <- ninetails::read_class_multiple(samples)
residue_data <- ninetails::read_residue_multiple(samples)

# 2. Annotate with gene symbols
class_data <- ninetails::annotate_with_biomart(class_data, species = "mmusculus")
residue_data <- ninetails::annotate_with_biomart(residue_data, species = "mmusculus")

# 3. Reclassify to remove artifacts
ninetails_data <- ninetails::reclassify_ninetails_data(
  residue_data = residue_data,
  class_data = class_data,
  grouping_factor = "sample_name",
  transcript_column = "ensembl_transcript_id_short",
  ref = "mmusculus"
)

class_data <- ninetails_data[[1]]
residue_data <- ninetails_data[[2]]

# 4. Count classifications
class_counts <- ninetails::count_class(class_data, grouping_factor = "group")

# 5. Merge and summarize
merged <- ninetails::merge_nonA_tables(class_data, residue_data)
summary <- ninetails::summarize_nonA(merged, summary_factors = "group")

# 6. Statistical comparison
fisher <- ninetails::calculate_fisher(
  class_data,
  grouping_factor = "group",
  comparison = c("WT", "KO")
)

# 7. Explore interactively
ninetails::launch_signal_browser(
  config = "config.yml"
)
```


---

## Summary of postprocessing functions

| Function | Description |
|---|---|
| `read_class_single()` | Load a single read_classes file |
| `read_class_multiple()` | Load multiple read_classes files with metadata |
| `read_residue_single()` | Load a single nonadenosine_residues file |
| `read_residue_multiple()` | Load multiple nonadenosine_residues files with metadata |
| `annotate_with_biomart()` | Add gene symbols and transcript IDs via biomaRt |
| `reclassify_ninetails_data()` | Correct misclassifications from AT-rich ends |
| `correct_class_data()` | Apply corrections to class table |
| `correct_residue_data()` | Apply corrections to residue table |
| `correct_labels()` | Fix label inconsistencies |
| `merge_nonA_tables()` | Combine class and residue data (one row per read) |
| `spread_nonA_residues()` | Reshape merged table with per-position columns |
| `count_class()` | Count reads by class or comment code |
| `count_residues()` | Count residues by type |
| `count_nonA_abundance()` | Count reads with 1, 2, 3+ non-A residues |
| `summarize_nonA()` | Per-transcript summary statistics |
| `calculate_fisher()` | Fisher's exact test between conditions |
