---
title: "Data postprocessing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data postprocessing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette describes functions for loading, processing, and analyzing **ninetails** output data.

## Reading files into R

Data postprocessing requires output from the main ninetails pipelines (`check_tails_dorado_DRS()`, `check_tails_dorado_cDNA()`, or `check_tails_guppy()`).

### Reading single files

```{r read-single}
# Read read_classes output
class_path <- "/path/to/read_classes.tsv"
class_data <- ninetails::read_class_single(class_path)

# Read nonadenosine_residues output
residue_path <- "/path/to/nonadenosine_residues.tsv"
residue_data <- ninetails::read_residue_single(residue_path)
```

### Reading multiple files

For experiments with multiple samples, use `read_class_multiple()` and `read_residue_multiple()` with a metadata table:

```{r read-multiple}
# Define metadata table
samples_table <- data.frame(
  sample_name = c("sample_1", "sample_2", "sample_3", "sample_4"),
  group = c("control", "control", "treatment", "treatment"),
  class_path = c(
    "/path/to/sample_1/read_classes.tsv",
    "/path/to/sample_2/read_classes.tsv",
    "/path/to/sample_3/read_classes.tsv",
    "/path/to/sample_4/read_classes.tsv"
  ),
  residue_path = c(
    "/path/to/sample_1/nonadenosine_residues.tsv",
    "/path/to/sample_2/nonadenosine_residues.tsv",
    "/path/to/sample_3/nonadenosine_residues.tsv",
    "/path/to/sample_4/nonadenosine_residues.tsv"
  )
)

# Read all files at once
class_data <- ninetails::read_class_multiple(samples_table)
residue_data <- ninetails::read_residue_multiple(samples_table)
```

> **Note:** Required metadata columns: `sample_name`, `group`, `class_path`, `residue_path`

### Using a YAML configuration file

Alternatively, provide metadata in a configuration file:

```{r read-yaml}
config <- yaml::yaml.load_file("config.yml")
samples_table <- data.frame(t(sapply(config$samples, unlist)))
rownames(samples_table) <- NULL

class_data <- ninetails::read_class_multiple(samples_table)
residue_data <- ninetails::read_residue_multiple(samples_table)
```

Example `config.yml`:

```yaml
samples:
  sample_1:
    sample_name: sample_1
    group: control
    class_path: /path/to/sample_1/read_classes.tsv
    residue_path: /path/to/sample_1/nonadenosine_residues.tsv
  sample_2:
    sample_name: sample_2
    group: treatment
    class_path: /path/to/sample_2/read_classes.tsv
    residue_path: /path/to/sample_2/nonadenosine_residues.tsv
```

---

## Correcting classification

Sometimes nucleotides from AT-rich transcript ends are misidentified as poly(A) tails. Use `reclassify_ninetails_data()` to minimize segmentation artifacts:

```{r reclassify}
ninetails_data <- ninetails::reclassify_ninetails_data(
  residue_data = residue_data,
  class_data = class_data,
  grouping_factor = "sample_name",
  transcript_column = "ensembl_transcript_id_short",
  ref = "mmusculus"
)

# Retrieve corrected data frames
class_data <- ninetails_data[[1]]
residue_data <- ninetails_data[[2]]
```

> **Note:** Apply this function before further analysis.

### Supported species

- *Arabidopsis thaliana*
- *Homo sapiens*
- *Mus musculus*
- *Saccharomyces cerevisiae*
- *Caenorhabditis elegans*
- *Trypanosoma brucei*

See function documentation for other species.

---

## Counting and summarizing

### Count by class

```{r count-class}
# Simple counts
class_counts <- ninetails::count_class(class_data)

# Grouped by sample
class_counts <- ninetails::count_class(
  class_data,
  grouping_factor = "sample_name"
)

# Detailed (by comments code) vs simplified (by class)
counts_detailed <- ninetails::count_class(class_data, detailed = TRUE)
counts_simple <- ninetails::count_class(class_data, detailed = FALSE)
```

### Count residues

```{r count-residues}
residue_counts <- ninetails::count_residues(
  residue_data,
  grouping_factor = "sample_name"
)
```

---

## Merging results

Combine class and residue data into a single table (one row per read):

```{r merge}
merged_tables <- ninetails::merge_nonA_tables(
  class_data = class_data,
  residue_data = residue_data,
  pass_only = TRUE
)
```

The output includes a `nonA_residues` column summarizing all non-A positions per read (5' to 3', separated by ":").

---

## Summarizing by transcript

Generate summary statistics per transcript:

```{r summarize}
summarized <- ninetails::summarize_nonA(
  merged_nonA_tables = merged_tables,
  summary_factors = "group",
  transcript_id_column = "ensembl_transcript_id_short"
)
```

Output includes:

- **counts**: Number of reads (total and by non-A type)
- **hits**: Number of non-A occurrences (a read may have multiple)
- Mean and median poly(A) tail length by transcript

---

## Statistical analysis

### Fisher's exact test

Compare non-adenosine frequencies between conditions:

```{r fisher}
fisher_results <- ninetails::calculate_fisher(
  ninetails_data,
  grouping_factor = "condition",
  comparison = c("control", "treatment")
)
```

---

## Typical workflow

```{r workflow}
library(ninetails)

# 1. Load data from multiple samples
samples <- data.frame(
  sample_name = c("WT_1", "WT_2", "KO_1", "KO_2"),
  group = c("WT", "WT", "KO", "KO"),
  class_path = c("WT_1_classes.tsv", "WT_2_classes.tsv",
                 "KO_1_classes.tsv", "KO_2_classes.tsv"),
  residue_path = c("WT_1_residues.tsv", "WT_2_residues.tsv",
                   "KO_1_residues.tsv", "KO_2_residues.tsv")
)

class_data <- ninetails::read_class_multiple(samples)
residue_data <- ninetails::read_residue_multiple(samples)

# 2. Reclassify to remove artifacts
ninetails_data <- ninetails::reclassify_ninetails_data(
  residue_data = residue_data,
  class_data = class_data,
  grouping_factor = "sample_name",
  transcript_column = "ensembl_transcript_id_short",
  ref = "mmusculus"
)

class_data <- ninetails_data[[1]]
residue_data <- ninetails_data[[2]]

# 3. Count classifications
class_counts <- ninetails::count_class(class_data, grouping_factor = "group")

# 4. Merge and summarize
merged <- ninetails::merge_nonA_tables(class_data, residue_data)
summary <- ninetails::summarize_nonA(merged, summary_factors = "group")

# 5. Statistical comparison
fisher <- ninetails::calculate_fisher(
  class_data,
  grouping_factor = "group",
  comparison = c("WT", "KO")
)
```
