---
title: "Analysis dashboard (Shiny)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysis dashboard (Shiny)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Ninetails includes an interactive Shiny dashboard for exploring analysis
results. The dashboard provides six tabs for classification overview,
residue composition, poly(A) length distribution, raw signal visualization,
configurable report download, and package information.

Two launcher functions are available depending on the pipeline used:

| Launcher | Pipeline | Signal input |
|---|---|---|
| `launch_signal_browser()` | Dorado DRS | POD5 files |
| `launch_signal_browser_guppy()` | Guppy legacy | fast5 files |

Both launchers support the same two usage modes:

- **Single-sample mode** — provide file paths directly; all tabs are
  available when classification and residue data are supplied.
- **Multi-sample mode** — provide a YAML configuration file pointing to
  multiple samples; enables comparative analysis across samples and
  experimental groups.

## Requirements

The dashboard requires several optional packages. Install them before
first use:

```{r install-deps}
install.packages(c("shiny", "plotly", "htmltools", "DT", "base64enc"))

# For multi-sample mode (YAML config):
install.packages("yaml")

# For rug density plots (position distribution):
install.packages("cowplot")
```

The Signal Viewer tab of `launch_signal_browser()` additionally requires a
working Python environment with the `pod5` package, accessible via
`reticulate`. See the [installation guide](https://github.com/LRB-IIMCB/ninetails/wiki)
for details. `launch_signal_browser_guppy()` does not require Python.

---

## Dorado DRS dashboard

### Quick start

#### Single-sample mode

```{r launch-single}
ninetails::launch_signal_browser(
  summary_file = "/path/to/dorado_summary.txt",
  pod5_dir     = "/path/to/pod5/",
  class_file   = "/path/to/read_classes.txt",
  residue_file = "/path/to/nonadenosine_residues.txt"
)
```

All four arguments are optional. When `class_file` and `residue_file` are
provided, the Classification, Residues, and Poly(A) length tabs become
active. When `summary_file` and `pod5_dir` are provided, the Signal
Viewer tab enables raw signal visualization.

If only signal data paths are provided (no `class_file`), only the Signal
Viewer tab is functional:

```{r launch-signal-only}
ninetails::launch_signal_browser(
  summary_file = "/path/to/dorado_summary.txt",
  pod5_dir     = "/path/to/pod5/"
)
```

#### Multi-sample mode

```{r launch-multi}
ninetails::launch_signal_browser(
  config = "/path/to/config.yml"
)
```

### YAML configuration file

```yaml
samples:
  KO_rep1:
    sample_name: KO_rep1
    group: KO
    class_path: /path/to/KO_rep1/read_classes.txt
    residue_path: /path/to/KO_rep1/nonadenosine_residues.txt
    dorado_summary: /path/to/KO_rep1/dorado_summary.txt   # optional
    pod5_dir: /path/to/KO_rep1/pod5/                      # optional

  WT_rep1:
    sample_name: WT_rep1
    group: WT
    class_path: /path/to/WT_rep1/read_classes.txt
    residue_path: /path/to/WT_rep1/nonadenosine_residues.txt
    dorado_summary: /path/to/WT_rep1/dorado_summary.txt
    pod5_dir: /path/to/WT_rep1/pod5/
```

| Field | Required | Description |
|---|---|---|
| `sample_name` | Yes | Display name used in plots and legends |
| `group` | Yes | Experimental condition for grouping |
| `class_path` | Yes | Path to `read_classes.txt` from ninetails |
| `residue_path` | Yes | Path to `nonadenosine_residues.txt` from ninetails |
| `dorado_summary` | No | Path to Dorado summary file (for Signal Viewer) |
| `pod5_dir` | No | Path to POD5 file directory (for Signal Viewer) |

### Launcher reference

```{r launcher-help-dorado, eval = FALSE}
?ninetails::launch_signal_browser
```

| Argument | Type | Description |
|---|---|---|
| `config` | character | Path to YAML config (multi-sample mode) |
| `summary_file` | character | Path to Dorado summary file |
| `pod5_dir` | character | Path to POD5 directory |
| `class_file` | character | Path to `read_classes.txt` |
| `residue_file` | character | Path to `nonadenosine_residues.txt` |
| `...` | | Additional arguments passed to `shiny::runApp()` |

---

## Guppy legacy dashboard

### Quick start

#### Single-sample mode

```{r launch-guppy-single}
ninetails::launch_signal_browser_guppy(
  nanopolish_file         = "/path/to/nanopolish_output.tsv",
  sequencing_summary_file = "/path/to/sequencing_summary.txt",
  workspace               = "/path/to/fast5/",
  class_file              = "/path/to/read_classes.txt",
  residue_file            = "/path/to/nonadenosine_residues.txt"
)
```

`nanopolish_file`, `sequencing_summary_file`, and `workspace` are needed
for the Signal Viewer tab. `class_file` and `residue_file` activate all
other analysis tabs. All arguments are optional.

#### Multi-sample mode

```{r launch-guppy-multi}
ninetails::launch_signal_browser_guppy(
  config = "/path/to/config_guppy.yml"
)
```

### YAML configuration file

```yaml
samples:
  KO_rep1:
    sample_name: KO_rep1
    group: KO
    class_path: /path/to/KO_rep1/read_classes.txt
    residue_path: /path/to/KO_rep1/nonadenosine_residues.txt
    polya_path: /path/to/KO_rep1/nanopolish_output.tsv   # optional
    seq_summary: /path/to/KO_rep1/sequencing_summary.txt # optional
    workspace: /path/to/KO_rep1/fast5/                   # optional

  WT_rep1:
    sample_name: WT_rep1
    group: WT
    class_path: /path/to/WT_rep1/read_classes.txt
    residue_path: /path/to/WT_rep1/nonadenosine_residues.txt
```

| Field | Required | Description |
|---|---|---|
| `sample_name` | Yes | Display name used in plots and legends |
| `group` | Yes | Experimental condition for grouping |
| `class_path` | Yes | Path to `read_classes.txt` from ninetails |
| `residue_path` | Yes | Path to `nonadenosine_residues.txt` from ninetails |
| `polya_path` | No | Path to Nanopolish polya output (for Signal Viewer) |
| `seq_summary` | No | Path to Guppy sequencing summary (for Signal Viewer) |
| `workspace` | No | Path to multi-fast5 directory (for Signal Viewer) |

### Guppy-specific dashboard features

The Guppy dashboard provides the same six tabs as the Dorado version, with
two additions:

**Nanopolish QC** (Classification tab) — an additional plot showing the
distribution of Nanopolish QC tags (PASS, ADAPTER, NOREGION, SUFFCLIP,
etc.) produced by `plot_nanopolish_qc()`.

**Fast5 Signal Viewer** — uses `plot_squiggle_fast5()` and
`plot_tail_range_fast5()` instead of POD5-based functions. Includes a
**Moves** toggle to show or hide basecaller move transitions in the signal
background. No Python dependency is required.

### Launcher reference

```{r launcher-help-guppy, eval = FALSE}
?ninetails::launch_signal_browser_guppy
```

| Argument | Type | Description |
|---|---|---|
| `config` | character | Path to YAML config (multi-sample mode) |
| `nanopolish_file` | character | Path to Nanopolish polya output |
| `sequencing_summary_file` | character | Path to Guppy sequencing summary |
| `workspace` | character | Path to multi-fast5 directory |
| `class_file` | character | Path to `read_classes.txt` |
| `residue_file` | character | Path to `nonadenosine_residues.txt` |
| `basecall_group` | character | Fast5 hierarchy level (default: `"Basecall_1D_000"`) |
| `...` | | Additional arguments passed to `shiny::runApp()` |

---

## Dashboard tabs

All tabs described below are available in both launchers unless noted
otherwise.

### Classification

The first tab provides an overview of the analysis results.

**Value boxes** at the top display five metrics: number of samples
analyzed, number of transcripts found, total read count, number of blank
reads, and number of decorated reads.

**Read Classification plot** shows the distribution of read categories
(decorated, blank, unclassified). Three views are available via the
"Classification view" dropdown:

- **Summary (N)** — simplified view with main categories
- **Detailed (R)** — expanded view with all comment codes
  (YAY, MPU, MAU, IRL, BAC, UNM)
- **Decorated (A)** — decorated reads only

**Non-A Abundance plot** shows the frequency of reads containing one, two,
or three or more separate non-adenosine residues per read
(`plot_nonA_abundance()`).

**Nanopolish QC plot** *(Guppy dashboard only)* — shows the distribution
of Nanopolish QC tags (`plot_nanopolish_qc()`).

**Sidebar controls:**

- *Grouping variable* — choose between `sample_name` or `group`
- *Filter by transcript* — server-side searchable dropdown
- *Show as frequency* — toggle between raw counts and proportions
- *Show descriptions* — toggle explanatory text below each plot

### Residues

The Residues tab provides three views of non-adenosine residue composition.

**Residue Counts plot** shows the distribution of C, G, and U residues
(`plot_residue_counts()`).

**Non-A Position Distribution** displays rug density plots showing where
non-adenosine residues are located along the poly(A) tail relative to
tail length (`plot_rug_density()`). Each residue type is subsampled to a
maximum of 1,000 points per sample.

**Summary Table** (when merged data is available) provides per-transcript
statistics (`summarize_nonA()`).

### Poly(A) length

Shows poly(A) tail length distributions across samples or groups
(`plot_tail_distribution()`). Controls include grouping, transcript filter,
condition selection, central tendency overlay, x-axis limit, normalization,
and color palette.

### Signal Viewer

Provides interactive visualization of raw nanopore signals.

**Dorado dashboard** — reads signals from POD5 files. Requires
`summary_file` and `pod5_dir`.

**Guppy dashboard** — reads signals from multi-fast5 files. Requires
`nanopolish_file`, `sequencing_summary_file`, and `workspace`. Includes a
**Moves** toggle to show basecaller move transitions.

Both versions provide:

- *Static Viewer* — two ggplot2 plots (entire signal and poly(A) region
  with non-A modification overlays)
- *Dynamic Explorer* — interactive Plotly chart with zoom, pan, and hover

### Download

Configurable report generator producing a self-contained HTML file with
all plots embedded as base64-encoded PNG images.

### About

Displays the ninetails logo, citation, links to GitHub, Wiki, pkgdown,
and Zenodo, and institute information.

---

## Annotated data

When the input data has been annotated with `annotate_with_biomart()`
(adding `symbol` and Ensembl transcript ID columns), the dashboard
automatically uses `symbol` as the transcript label in all "Filter by
transcript" dropdowns.

```{r annotate}
class_data_annotated <- ninetails::annotate_with_biomart(
  class_data,
  species = "mmusculus"
)

residue_data_annotated <- ninetails::annotate_with_biomart(
  residue_data,
  species = "mmusculus"
)
```

---

## Running on a server

### Custom port and host

Pass `port` and `host` through `...` to make the app accessible on your
network:

```{r runapp-args}
# Dorado
ninetails::launch_signal_browser(
  config = "config.yml",
  port   = 8080,
  host   = "0.0.0.0"
)

# Guppy
ninetails::launch_signal_browser_guppy(
  config = "config_guppy.yml",
  port   = 8080,
  host   = "0.0.0.0"
)
```

### Sourcing the app directly

The app can also be sourced directly from the installed package without
using the launcher function. This is useful for embedding in scripts or
for Shiny Server deployments where `shinyOptions` must be set manually:

```{r source-app}
library(ninetails)
library(shiny)

# Set data via shinyOptions before sourcing
shiny::shinyOptions(
  ninetails.class_data    = class_data,
  ninetails.residue_data  = residue_data,
  ninetails.merged_data   = merged_data,
  ninetails.signal_config = list(),
  ninetails.summary_file  = "/path/to/dorado_summary.txt",
  ninetails.pod5_dir      = "/path/to/pod5/"
)

# Source the app into an isolated environment
app_dir <- system.file("app", package = "ninetails")       # Dorado
# app_dir <- system.file("app_guppy", package = "ninetails")  # Guppy

app_env <- new.env(parent = globalenv())
source(file.path(app_dir, "app.R"), local = app_env)

shiny::shinyApp(ui = app_env$ui, server = app_env$server)
```

### Deploying to Shiny Server

A unified deployment wrapper is included in the package at
`inst/deployment/app.R`. It auto-detects the pipeline (Dorado or Guppy)
from the YAML configuration and sources the correct app at runtime.

#### Setup

```{r deploy, eval = FALSE}
# 1. Create the app directory on the server
deploy_dir <- "/srv/shiny-server/ninetails"
dir.create(deploy_dir, recursive = TRUE, showWarnings = FALSE)

# 2. Copy the deployment wrapper
file.copy(
  system.file("deployment", "app.R", package = "ninetails"),
  file.path(deploy_dir, "app.R"),
  overwrite = TRUE
)

# 3. Copy your YAML configuration
file.copy("config.yml", file.path(deploy_dir, "config.yml"))

# 4. Copy or symlink static assets
# For Dorado app:
file.symlink(
  system.file("app", "www", package = "ninetails"),
  file.path(deploy_dir, "www")
)
# For Guppy app (if config describes Guppy data):
# file.symlink(
#   system.file("app_guppy", "www", package = "ninetails"),
#   file.path(deploy_dir, "www")
# )
```

#### Configuring the deployment wrapper

Edit the top section of the deployed `app.R` to set three variables:

```r
# In /srv/shiny-server/ninetails/app.R:

# Path to the YAML config (relative to app.R)
config_path <- file.path(getwd(), "config.yml")

# Python binary with 'pod5' installed (Dorado Signal Viewer only)
# Set to NULL if signal visualization is not needed
PYTHON_PATH <- "/home/user/miniconda3/envs/r-reticulate/bin/python"

# Fast5 basecall group (Guppy Signal Viewer only)
BASECALL_GROUP <- "Basecall_1D_000"
```

The wrapper auto-detects the pipeline from the YAML: if the first sample
contains `pod5_dir` or `dorado_summary` fields, the Dorado app is used;
if it contains `workspace` or `polya_path`, the Guppy app is used.

Common Python path examples:

| Environment | Typical path |
|---|---|
| System Python | `/usr/bin/python3` |
| Conda env | `/home/user/miniconda3/envs/r-reticulate/bin/python` |
| virtualenv | `/home/user/.virtualenvs/ninetails/bin/python` |
| pyenv | `/home/user/.pyenv/versions/3.11.0/bin/python` |

Run `which python` in the environment where `pod5` is installed to find
the correct path.

#### Directory structure

```
/srv/shiny-server/ninetails/
├── app.R          # deployment wrapper (from inst/deployment/)
├── config.yml     # your YAML configuration
└── www/           # symlink to inst/app/www/ or inst/app_guppy/www/
    ├── logo.png
    ├── favicon.ico
    └── IIMCB_logo.png
```

#### Requirements

- ninetails and all Suggests dependencies must be installed system-wide
  (accessible to the `shiny` user)
- All data file paths in `config.yml` must be readable by the `shiny` user
- **Dorado Signal Viewer**: `PYTHON_PATH` must be set; the specified
  environment must have `pod5` installed (`pip install pod5`); the
  `shiny` user must have execute permissions on the Python binary
- **Guppy Signal Viewer**: no Python dependency; the `shiny` user must
  have read access to the fast5 files and to the Nanopolish output

Restart Shiny Server after placing the files:

```bash
sudo systemctl restart shiny-server
```

The dashboard will be accessible at `http://your-server:3838/ninetails/`.

> **Note:** When you update the ninetails package, the deployed app picks
> up changes automatically (it sources from the installed package at
> runtime). To change the dataset, edit `config.yml` and restart Shiny
> Server.

---

## Static assets

The dashboard expects three image files in `inst/app/www/` (Dorado) or
`inst/app_guppy/www/` (Guppy):

- `logo.png` — ninetails logo (copy from `man/figures/logo.png`)
- `favicon.ico` — browser tab icon (copy from `pkgdown/favicon/`)
- `IIMCB_logo.png` — IIMCB institute logo (displayed in the About tab)

---

## Troubleshooting

**Rug density plots are empty or show an error**
`plot_rug_density()` requires the `cowplot` package. Install it with
`install.packages("cowplot")`.

**Download button produces an error**
The report generation requires `base64enc`. Install with
`install.packages("base64enc")`.

**Signal Viewer shows "No POD5 file found"** *(Dorado)*
Ensure `pod5_dir` points to the directory containing `.pod5` files. The
app searches recursively within that directory.

**Signal Viewer shows "Extraction failed"** *(Dorado)*
The Python `pod5` package must be installed and accessible via
`reticulate`. Check with:

```{r check-pod5}
reticulate::py_module_available("pod5")
```

**Signal Viewer shows no signal** *(Guppy)*
Ensure `nanopolish_file`, `sequencing_summary_file`, and `workspace` all
point to valid paths accessible to the R session. The read ID selected
must exist in the Nanopolish output.

**Classification or Residue plots are blank**
In single-sample mode, `class_file` and/or `residue_file` must be
provided. Without them, the corresponding tabs show placeholder messages.

**"Column `more` doesn't exist" in Non-A Abundance plot**
This occurs when no reads have 3 or more non-A residues. Update
`plot_nonA_abundance()` to the latest version which handles missing
columns after pivot.

**Y-axis shows scientific notation**
The dashboard disables scientific notation globally with
`options(scipen = 999)` at startup. If you still see scientific notation,
ensure you are running the latest version of the app.
