---
title: "Training dataset preparation for the cDNA models"
author: "Natalia Guminska"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Configuration

```{r, load libs, message = FALSE, include=FALSE}

library(vroom)
library(reticulate)
library(dplyr)
library(ggplot2)
library(ninetails)
library(keras)
library(tensorflow)

```


```{r message = FALSE, echo=FALSE}
# use virtualenv to communicate ninetails with correct python version
# depends on configuration; the pod5 module must be installed there
reticulate::use_condaenv("r-reticulate")

# number of cores for processing
num_cores <- 30

# output directory for intermediate objects
out_dir <- "/home/nagumi/CNN_spikes_cdna"
if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

```


# Rationale

The cDNA pipeline needs two models, one per read orientation, because the
two orientations see different signals:

- polyA reads (sense strand): transcript body | poly(A) tail | VNP (rc). The
  tail signal runs from the transcript body towards the 3' end, so it is
  reversed with respect to the DRS convention (3' end first). Residues are
  observed as themselves (C, G, T in a dA background).
- polyT reads (antisense strand): VNP | poly(T) tail | transcript body (rc).
  The tail signal runs from the 3' end of the original tail towards the body
  and every residue is observed as its complement (C -> G, G -> C, U -> A) in
  a dT background.

No signal reversal is applied anywhere (the production pipeline feeds the
signals as extracted, so the training data must match). Class labels always
refer to the residue in the original RNA tail, known from the experimental
design of the spike-in constructs, not to the base actually read in polyT
orientation.

The polarity of the signal deviation (peak or valley) caused by each residue
was hardcoded in the DRS routine (G = peak, C/U = valley). It is unknown for
the DNA chemistry and may differ between orientations, therefore it is
established here from the labelled data (lookup 2) before the chunks are
filtered.


# Input definition

One entry per labelled dataset. Each construct carries a single residue type
inserted into the tail (or none, for the A control). The `contig` field selects
the reads mapping to the construct of interest, so reads from other references
present in the same run are excluded. Use the same name for the contig as in
the `alignment_genome` column of the Dorado summary.

```{r inputs}
spikein_inputs <- list(
  C = list(
    dorado_summary = "/path/to/C_Rluc_cdna/dorado_summary.txt",
    bam_file       = "/path/to/C_Rluc_cdna/aligned.bam",
    pod5_dir       = "/path/to/C_Rluc_cdna/pod5/",
    contig         = "RlucB"
  ),
  G = list(
    dorado_summary = "/path/to/G_Rluc_cdna/dorado_summary.txt",
    bam_file       = "/path/to/G_Rluc_cdna/aligned.bam",
    pod5_dir       = "/path/to/G_Rluc_cdna/pod5/",
    contig         = "Rluc-sh"
  ),
  U = list(
    dorado_summary = "/path/to/U_Rluc_cdna/dorado_summary.txt",
    bam_file       = "/path/to/U_Rluc_cdna/aligned.bam",
    pod5_dir       = "/path/to/U_Rluc_cdna/pod5/",
    contig         = "RlucA"
  ),
  A = list(
    dorado_summary = "/path/to/A_Rluc_cdna/dorado_summary.txt",
    bam_file       = "/path/to/A_Rluc_cdna/aligned.bam",
    pod5_dir       = "/path/to/A_Rluc_cdna/pod5/",
    contig         = "RlucD"
  )
)

nucleotides <- names(spikein_inputs)
tail_types <- c("polyA", "polyT")
```


# Step 1: signal extraction and orientation split

Each class is processed separately. The function filters the summary
(mapped, MAPQ > 0, valid coordinates, tail >= 10 nt), keeps the selected
contig, classifies the orientation from the basecalled sequences and
extracts the winsorized, downsampled (20%) tail signals from POD5.

```{r extract signals}
extracted <- list()

for (nt in nucleotides) {
  cat("\n=== ", nt, " ===\n")
  extracted[[nt]] <- ninetails::extract_tail_signals_trainingset_cdna(
    dorado_summary = spikein_inputs[[nt]]$dorado_summary,
    bam_file = spikein_inputs[[nt]]$bam_file,
    pod5_dir = spikein_inputs[[nt]]$pod5_dir,
    num_cores = num_cores,
    contig = spikein_inputs[[nt]]$contig
  )
}

save(extracted, file = file.path(out_dir, "extracted_signals_cdna.RData"))
```


## Lookup 1: are the reads what they should be?

Every read in the annotation must map to the expected contig, and both
orientations should be reasonably populated. A high share of unidentified
reads points to truncated reads or wrong primers. The signal length after
downsampling should scale with the tail length (roughly 0.2 x samples per nt
x tail length); reads far off the trend deserve a look.

```{r lookup 1 orientation and contig}
for (nt in nucleotides) {
  cat("\n=== ", nt, " ===\n")
  annotation <- extracted[[nt]]$read_annotation
  print(table(contig = annotation$contig, tail_type = annotation$tail_type))
  cat("\npoly(A) length summary by orientation:\n")
  print(tapply(annotation$poly_tail_length, annotation$tail_type, summary))
  cat("\nsignal length (data points) summary by orientation:\n")
  print(tapply(annotation$signal_length, annotation$tail_type, summary))
}
```


```{r lookup 1 tail length vs signal length, fig.width=10, fig.height=4}
annotation_all <- dplyr::bind_rows(
  lapply(nucleotides, function(nt) {
    dplyr::mutate(extracted[[nt]]$read_annotation, nucleotide = nt)
  })
)

length_check <- ggplot2::ggplot(
  data = annotation_all[annotation_all$tail_type != "unidentified", ],
  ggplot2::aes(x = poly_tail_length, y = signal_length, color = tail_type)
) +
  ggplot2::geom_point(alpha = .3, size = .8) +
  ggplot2::facet_wrap(~nucleotide, ncol = 4) +
  ggplot2::scale_color_manual(values = c("polyA" = "#f56042", "polyT" = "#089bcc")) +
  ggplot2::labs(
    title = "Downsampled signal length versus Dorado tail length",
    x = "\npoly(A) length [nt]",
    y = "signal length [data points]\n"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "top", legend.title = ggplot2::element_blank())

print(length_check)
```


# Step 2: feature lists (pseudomoves) per class and orientation

The retention criterion follows the original routine: C/G/U reads must
contain at least one non-zero pseudomove run of length >= 4, A reads must
contain none.

```{r feature lists}
features <- list()

for (nt in nucleotides) {
  for (tt in tail_types) {
    cat("\n=== ", nt, tt, " ===\n")
    signal_list <- if (tt == "polyA") {
      extracted[[nt]]$polya_signals
    } else {
      extracted[[nt]]$polyt_signals
    }
    if (length(signal_list) == 0) {
      cat("no reads, skipping\n")
      next
    }
    features[[nt]][[tt]] <- ninetails::create_tail_feature_list_trainingset_cdna(
      signal_list = signal_list,
      num_cores = num_cores,
      nucleotide = nt
    )
    cat(sprintf(
      "retained %d reads, discarded %d\n",
      length(features[[nt]][[tt]]$tail_feature_list),
      length(features[[nt]][[tt]]$discarded_readnames)
    ))
  }
}

save(features, file = file.path(out_dir, "features_cdna.RData"))
```


## Lookup 2: pseudomove polarity

For every labelled class and orientation, count the peak (+1) and valley
(-1) runs. The dominant polarity is the `value` passed to
`filter_nonA_chunks_trainingset()` in step 4. Expect one polarity to
dominate clearly; if the two columns are comparable, the residue-induced
distortion is not captured by the z-score filter in this orientation and
the data should be inspected before proceeding. The A class is listed only
as a sanity check (its retained reads have no qualifying runs by
construction).

```{r lookup 2 polarity table}
polarity_table <- dplyr::bind_rows(
  lapply(nucleotides, function(nt) {
    dplyr::bind_rows(lapply(names(features[[nt]]), function(tt) {
      run_counts <- ninetails::count_pseudomove_runs_trainingset_cdna(
        features[[nt]][[tt]]
      )
      data.frame(
        nucleotide = nt,
        tail_type = tt,
        reads = nrow(run_counts),
        peak_runs = sum(run_counts$peak_runs),
        valley_runs = sum(run_counts$valley_runs),
        reads_with_peak_only = sum(run_counts$peak_runs > 0 & run_counts$valley_runs == 0),
        reads_with_valley_only = sum(run_counts$valley_runs > 0 & run_counts$peak_runs == 0),
        reads_with_both = sum(run_counts$peak_runs > 0 & run_counts$valley_runs > 0),
        stringsAsFactors = FALSE
      )
    }))
  })
)

print(polarity_table)
```


```{r lookup 2 polarity decision}
# polarity per class and orientation, read off the table above
# fill in after inspection: 1 = peak, -1 = valley
polarity <- list(
  C = c(polyA = NA, polyT = NA),
  G = c(polyA = NA, polyT = NA),
  U = c(polyA = NA, polyT = NA)
)

# suggested values (dominant polarity); verify visually before accepting
for (nt in c("C", "G", "U")) {
  for (tt in names(features[[nt]])) {
    row <- polarity_table[polarity_table$nucleotide == nt & polarity_table$tail_type == tt, ]
    polarity[[nt]][[tt]] <- ifelse(row$peak_runs >= row$valley_runs, 1, -1)
  }
}

print(polarity)
```


## Lookup 2b: visual inspection of single reads

A few randomly drawn reads per class and orientation, with the qualifying
pseudomove runs highlighted. The highlighted regions are the future chunk
centres. Genuine residue-induced distortions are short, localised
excursions from the flat homopolymer baseline; extended noisy stretches or
runs at the very ends indicate segmentation errors.

```{r lookup 2b example reads, fig.width=15, fig.height=4}
set.seed(123)

for (nt in nucleotides) {
  for (tt in names(features[[nt]])) {
    readnames <- names(features[[nt]][[tt]]$tail_feature_list)
    if (length(readnames) == 0) {
      next
    }
    for (rn in sample(readnames, min(3, length(readnames)))) {
      example <- ninetails::plot_tail_features_trainingset_cdna(
        readname = rn,
        tail_feature_list = features[[nt]][[tt]]
      ) +
        ggplot2::labs(title = paste0(nt, " / ", tt, " / read ", rn))
      print(example)
    }
  }
}
```


# Step 3: chunk extraction

Reuses the Guppy training-set chunkers: modification-centred 100-point
chunks for C/G/U, overlapping 100-point windows (50-point overlap) for A.

```{r chunks}
chunks <- list()

for (nt in nucleotides) {
  for (tt in names(features[[nt]])) {
    cat("\n=== ", nt, tt, " ===\n")
    if (nt == "A") {
      chunks[[nt]][[tt]] <- ninetails::create_tail_chunk_list_A(
        features[[nt]][[tt]],
        num_cores
      )
    } else {
      chunks[[nt]][[tt]] <- ninetails::create_tail_chunk_list_trainingset(
        features[[nt]][[tt]],
        num_cores
      )
    }
  }
}

save(chunks, file = file.path(out_dir, "chunks_cdna.RData"))
```


## Lookup 3: chunk structure

All chunks must be exactly 100 data points long and contain no NAs (the
GAF transformation requires it). The number of chunks per read is reported
per class and orientation.

```{r lookup 3 chunk structure}
for (nt in nucleotides) {
  for (tt in names(chunks[[nt]])) {
    chunk_list <- chunks[[nt]][[tt]]
    if (nt == "A") {
      chunk_lengths <- unlist(lapply(chunk_list, function(x) sapply(x, length)))
      chunk_nas <- unlist(lapply(chunk_list, function(x) sapply(x, function(y) any(is.na(y)))))
    } else {
      chunk_lengths <- unlist(lapply(chunk_list, function(x) sapply(x, function(y) length(y$chunk_sequence))))
      chunk_nas <- unlist(lapply(chunk_list, function(x) sapply(x, function(y) any(is.na(y$chunk_sequence)))))
    }
    cat(sprintf(
      "%s / %s: %d reads, %d chunks, all length 100: %s, no NAs: %s\n",
      nt, tt,
      length(chunk_list),
      length(chunk_lengths),
      all(chunk_lengths == 100),
      !any(chunk_nas)
    ))
    print(summary(sapply(chunk_list, length)))
  }
}
```


```{r lookup 3 example chunk, fig.width=8, fig.height=3}
set.seed(123)

# one random modification-centred chunk per class and orientation
for (nt in c("C", "G", "U")) {
  for (tt in names(chunks[[nt]])) {
    chunk_list <- chunks[[nt]][[tt]]
    readname <- sample(names(chunk_list), 1)
    chunk_name <- sample(names(chunk_list[[readname]]), 1)
    example_chunk <- ninetails::plot_tail_chunk(
      chunk_name = chunk_name,
      tail_chunk_list = chunk_list
    ) +
      ggplot2::labs(title = paste0(nt, " / ", tt, " / chunk ", chunk_name))
    print(example_chunk)
  }
}
```


# Step 4: filtering of residue-containing chunks (C, G, U)

Chunks are retained if their pseudomoves contain a run of the established
polarity (>= 4) and their start position is non-negative (no left padding).

```{r filter chunks}
filtered_chunks <- list()

for (nt in c("C", "G", "U")) {
  for (tt in names(chunks[[nt]])) {
    cat("\n=== ", nt, tt, " (value = ", polarity[[nt]][[tt]], ") ===\n")
    if (is.na(polarity[[nt]][[tt]])) {
      stop("Polarity not established for ", nt, " / ", tt, ". See lookup 2.")
    }
    filtered_chunks[[nt]][[tt]] <- ninetails::filter_nonA_chunks_trainingset(
      chunks[[nt]][[tt]],
      value = polarity[[nt]][[tt]],
      num_cores = num_cores
    )
  }
}

# A chunks are not filtered
filtered_chunks[["A"]] <- chunks[["A"]]

save(filtered_chunks, file = file.path(out_dir, "filtered_chunks_cdna.RData"))
```


## Lookup 4: retained fraction after filtering

A very low retained fraction (a few percent) suggests the polarity was
chosen wrongly or the distortions are too weak to be captured.

```{r lookup 4 retained fraction}
for (nt in c("C", "G", "U")) {
  for (tt in names(chunks[[nt]])) {
    n_before <- sum(sapply(chunks[[nt]][[tt]], length))
    n_after <- sum(sapply(filtered_chunks[[nt]][[tt]], length))
    cat(sprintf(
      "%s / %s: %d -> %d chunks retained (%.1f%%), %d -> %d reads\n",
      nt, tt, n_before, n_after, 100 * n_after / max(n_before, 1),
      length(chunks[[nt]][[tt]]), length(filtered_chunks[[nt]][[tt]])
    ))
  }
}
```


# Step 5: GASF + GADF datasets

```{r gafs}
gafs <- list()

for (nt in nucleotides) {
  for (tt in names(filtered_chunks[[nt]])) {
    cat("\n=== ", nt, tt, " ===\n")
    if (nt == "A") {
      gafs[[nt]][[tt]] <- ninetails::create_gaf_list_A(filtered_chunks[[nt]][[tt]], num_cores)
    } else {
      gafs[[nt]][[tt]] <- ninetails::create_gaf_list(filtered_chunks[[nt]][[tt]], num_cores)
    }
  }
}

save(gafs, file = file.path(out_dir, "GADF_GASF_cdna.RData"))
```


## Lookup 5: GAF structure and class sizes

Every element must be a 100 x 100 x 2 array with values in [0, 1] and a
unique `<readname>_<index>` name. The class sizes per orientation decide the
sampling size in step 6.

```{r lookup 5 gaf structure}
gaf_sizes <- list()

for (nt in nucleotides) {
  for (tt in names(gafs[[nt]])) {
    gaf_list <- gafs[[nt]][[tt]]
    dims_ok <- all(sapply(gaf_list, function(x) identical(dim(x), c(100L, 100L, 2L))))
    range_ok <- all(sapply(gaf_list, function(x) min(x) >= 0 && max(x) <= 1))
    names_ok <- !is.null(names(gaf_list)) && !any(duplicated(names(gaf_list)))
    gaf_sizes[[tt]][[nt]] <- length(gaf_list)
    cat(sprintf(
      "%s / %s: %d GAFs, dims ok: %s, range ok: %s, unique names: %s\n",
      nt, tt, length(gaf_list), dims_ok, range_ok, names_ok
    ))
  }
}

print(gaf_sizes)
```


```{r lookup 5 example gaf, fig.width=4, fig.height=4}
set.seed(123)

for (nt in nucleotides) {
  for (tt in names(gafs[[nt]])) {
    gaf_name <- sample(names(gafs[[nt]][[tt]]), 1)
    example_gaf <- ninetails::plot_gaf(gaf_name = gaf_name, gaf_list = gafs[[nt]][[tt]])
    print(example_gaf)
  }
}
```


# Step 6 and 7: sampling, splitting and conversion to keras arrays

Same procedure as in the DRS routine, applied separately to each
orientation: equal number of GAFs per class (A, C, G, U), 90:10 split into
training and validation, shuffle, encoding A = 0, C = 1, G = 2, U = 3,
conversion to a 4-D array (n x 100 x 100 x 2).

```{r dataset builder}
# builds train and validation lists for one orientation
# gaf_by_class: named list (A, C, G, U) of GAF lists
build_cnn_dataset <- function(gaf_by_class, n_per_class, seed = 123456) {

  #setting seed for reprex
  set.seed(seed)

  # first, random sample the same number of GAFs from each class
  sampled <- lapply(gaf_by_class, function(x) sample(x, size = n_per_class))

  # extract names of each selected gafs group (needed for encoding)
  class_names <- lapply(sampled, names)

  # split dataset for training and validation data (90:10)
  n_train <- floor(0.9 * n_per_class)
  train_set <- lapply(sampled, function(x) x[1:n_train])
  valid_set <- lapply(sampled, function(x) x[(n_train + 1):n_per_class])

  # encoding: A = 0, C = 1, G = 2, U = 3
  encode_labels <- function(gaf_list) {
    ifelse(names(gaf_list) %in% class_names$C, 1,
           ifelse(names(gaf_list) %in% class_names$G, 2,
                  ifelse(names(gaf_list) %in% class_names$U, 3, 0)))
  }

  # merge, shuffle and encode
  to_cnn_input <- function(set_list) {
    merged <- c(set_list$A, set_list$C, set_list$G, set_list$U)
    merged <- sample(merged, size = length(merged), replace = FALSE)
    labels <- encode_labels(merged)
    chunk_names <- names(merged)
    # transform into keras array
    names(merged) <- NULL
    merged <- simplify2array(merged)
    merged <- aperm(merged, c(4, 1, 2, 3))
    out <- list()
    out[["x"]] <- merged
    out[["y"]] <- labels
    out[["chunk_names"]] <- chunk_names
    return(out)
  }

  dataset <- list()
  dataset[["train"]] <- to_cnn_input(train_set)
  dataset[["validation"]] <- to_cnn_input(valid_set)
  return(dataset)
}
```


```{r build datasets}
cnn_datasets <- list()

for (tt in tail_types) {
  gaf_by_class <- lapply(nucleotides, function(nt) gafs[[nt]][[tt]])
  names(gaf_by_class) <- nucleotides
  if (any(sapply(gaf_by_class, is.null))) {
    cat(tt, ": missing class, skipping\n")
    next
  }
  # class size: the smallest class; lower it manually to prevent overfitting
  # (9440 per class was used for the DRS model)
  n_per_class <- min(sapply(gaf_by_class, length))
  cat(sprintf("%s: %d GAFs per class\n", tt, n_per_class))
  cnn_datasets[[tt]] <- build_cnn_dataset(gaf_by_class, n_per_class)
}

train_cnn_polyA <- cnn_datasets$polyA$train
validation_cnn_polyA <- cnn_datasets$polyA$validation
train_cnn_polyT <- cnn_datasets$polyT$train
validation_cnn_polyT <- cnn_datasets$polyT$validation

save(train_cnn_polyA, validation_cnn_polyA,
     file = file.path(out_dir, "cnn_dataset_polyA.RData"))
save(train_cnn_polyT, validation_cnn_polyT,
     file = file.path(out_dir, "cnn_dataset_polyT.RData"))

rm(gafs, filtered_chunks, chunks)
gc()
```


## Lookup 6: dataset sanity

Dimensions, class balance, label consistency (re-derived from the chunk
names against the stored labels), and read-level leakage between training
and validation. The original DRS routine split by chunk, so chunks of the
same read may land on both sides; the leakage figure quantifies it so the
decision to split by read instead can be taken knowingly.

```{r lookup 6 dataset sanity}
for (tt in names(cnn_datasets)) {
  for (part in c("train", "validation")) {
    dataset <- cnn_datasets[[tt]][[part]]
    cat(sprintf("\n=== %s / %s ===\n", tt, part))
    cat("x dims:", dim(dataset$x), "\n")
    cat("y length:", length(dataset$y), "\n")
    print(table(dataset$y))
    stopifnot(dim(dataset$x)[1] == length(dataset$y))
    stopifnot(length(dataset$chunk_names) == length(dataset$y))
    # label consistency: chunk names must belong to the class they are labelled with
    readnames <- sub("_[0-9]+$", "", dataset$chunk_names)
    class_of_read <- lapply(nucleotides, function(nt) {
      extracted[[nt]]$read_annotation$read_id
    })
    names(class_of_read) <- nucleotides
    expected <- ifelse(readnames %in% class_of_read$C, 1,
                       ifelse(readnames %in% class_of_read$G, 2,
                              ifelse(readnames %in% class_of_read$U, 3, 0)))
    cat("labels consistent with source dataset:", all(expected == dataset$y), "\n")
  }
  # read-level leakage between training and validation
  train_reads <- unique(sub("_[0-9]+$", "", cnn_datasets[[tt]]$train$chunk_names))
  valid_reads <- unique(sub("_[0-9]+$", "", cnn_datasets[[tt]]$validation$chunk_names))
  shared <- intersect(train_reads, valid_reads)
  cat(sprintf(
    "%s: %d of %d validation reads also contribute chunks to training (%.1f%%)\n",
    tt, length(shared), length(valid_reads), 100 * length(shared) / max(length(valid_reads), 1)
  ))
}
```


# Model training (one model per orientation)

Architecture of the DRS model (`gasf_gadf_combined_model_refiltered_accuracy93`).
The `chunk_names` element is dropped before fitting.

```{r model definition}
build_model <- function() {
  model <- keras::keras_model_sequential() %>%
    keras::layer_conv_2d(filters = 32, kernel_size = c(5, 5), activation = "linear",
                         input_shape = c(100, 100, 2), kernel_initializer = "he_normal",
                         padding = "valid") %>%
    keras::layer_batch_normalization() %>%
    keras::layer_activation_relu() %>%
    keras::layer_max_pooling_2d(pool_size = c(2, 2)) %>%
    keras::layer_dropout(0.5) %>%
    keras::layer_conv_2d(filters = 32, kernel_size = c(5, 5), activation = "linear",
                         kernel_initializer = "he_normal", padding = "valid") %>%
    keras::layer_batch_normalization() %>%
    keras::layer_activation_relu() %>%
    keras::layer_max_pooling_2d(pool_size = c(2, 2)) %>%
    keras::layer_dropout(0.5) %>%
    keras::layer_conv_2d(filters = 64, kernel_size = c(3, 3), activation = "linear",
                         kernel_initializer = "he_normal", padding = "same") %>%
    keras::layer_batch_normalization() %>%
    keras::layer_activation_relu() %>%
    keras::layer_max_pooling_2d(pool_size = c(2, 2)) %>%
    keras::layer_dropout(0.5) %>%
    keras::layer_conv_2d(filters = 128, kernel_size = c(3, 3), activation = "linear",
                         kernel_initializer = "he_normal", padding = "same") %>%
    keras::layer_batch_normalization() %>%
    keras::layer_activation_relu() %>%
    keras::layer_max_pooling_2d(pool_size = c(2, 2)) %>%
    keras::layer_dropout(0.5) %>%
    keras::layer_flatten() %>%
    keras::layer_dense(units = 4, activation = "softmax")

  model %>% keras::compile(
    optimizer = "adam",
    loss = "sparse_categorical_crossentropy",
    metrics = "accuracy"
  )
  return(model)
}

# print dot for every epoch to check that it is actually working (for debug purposes)
print_dot_callback <- keras::callback_lambda(
  on_epoch_end = function(epoch, logs) {
    if (epoch %% 80 == 0) cat("\n")
    cat(".")
  }
)

# define when to stop the training if validation loss stops decreasing
early_stop <- keras::callback_early_stopping(monitor = "val_loss", patience = 3)

#define how many epochs use to train this thing
epochs <- 20
```


```{r train polyA}
model_polyA <- build_model()

history_polyA <- model_polyA %>% keras::fit(
  x = train_cnn_polyA$x,
  y = train_cnn_polyA$y,
  epochs = epochs,
  validation_split = 0.2,
  batch_size = 50,
  callbacks = list(early_stop, print_dot_callback,
                   keras::callback_reduce_lr_on_plateau(factor = 0.05))
)

plot(history_polyA)

keras::save_model_hdf5(
  model_polyA,
  file.path(out_dir, paste0("gasf_gadf_cdna_polyA_model_", format(Sys.Date(), "%Y%m%d"), ".h5")),
  include_optimizer = TRUE
)
```


```{r train polyT}
model_polyT <- build_model()

history_polyT <- model_polyT %>% keras::fit(
  x = train_cnn_polyT$x,
  y = train_cnn_polyT$y,
  epochs = epochs,
  validation_split = 0.2,
  batch_size = 50,
  callbacks = list(early_stop, print_dot_callback,
                   keras::callback_reduce_lr_on_plateau(factor = 0.05))
)

plot(history_polyT)

keras::save_model_hdf5(
  model_polyT,
  file.path(out_dir, paste0("gasf_gadf_cdna_polyT_model_", format(Sys.Date(), "%Y%m%d"), ".h5")),
  include_optimizer = TRUE
)
```


## Lookup 7: validation confusion matrices

```{r validation}
library(caret)

for (tt in names(cnn_datasets)) {
  model <- if (tt == "polyA") model_polyA else model_polyT
  validation <- cnn_datasets[[tt]]$validation
  test_prediction <- model %>% predict(validation$x) %>% keras::k_argmax()
  test_prediction <- as.numeric(test_prediction)
  cat("\n=== ", tt, " ===\n")
  print(caret::confusionMatrix(as.factor(test_prediction), as.factor(validation$y)))
}
```
